---
title: "Stat 133 HW04: String Manipulation and Regex"
author: Zhiqiang Liao, 24229382
output: pdf_document
---

## Introduction

This assignment has two purposes:

a) to familiarize you with manipulating character strings
b) to introduce you to regular expressions in R

Submit your assignment to bcourses, specifically turn in your __Rmd__ (R markdown) file as well as the produced pdf file. Make sure to change the argument ```eval=TRUE``` inside every testing code chunk.


-----

## Names of Files

Imagine that you need to generate the names of 4 data files (with .csv extension). All the files have the same prefix name but each of them has a different number: ```file01.csv```, ```file02.csv```, ```file03.csv```, and ```file04.csv```. We can generate a character vector with these names in R. One naive solution would be to write something like this:
```{r}
files <- c('file01.csv', 'file02.csv', 'file03.csv', 'file04.csv')
```

Now imagine that you need to generate 100 file names. You could write a vector with 100 file names but it's going to take you a while.

How would you generate the corresponding character vector ```files``` in R containing 100 files names:  ```file01.csv, file02.csv, file03.csv, ..., file99.csv, file100.csv```? Notice that the numbers of the first 9 files start with 0.

```{r}
# vector of file names
library(stringr)
sprintf("file%s.csv", str_pad(c(1:100), 2, "left", "0"))
```

-----

## USA States Names 

One of the datasets that come in R is ```USArrests```. The row names of this data correspond to the 50 states. We can create a vector ```states``` with the row names:
```{r}
states <- rownames(USArrests)
head(states, n = 5)
```

Use ```nchar()``` to answer the following questions:

- Obtain a frequency table with the number of characters of the states' names
- What are the states with the longest names?
- What are the states with the shortest names?
- What's the most common length of names (i.e. the mode)?

```{r}
# your answers
table(nchar(states))
states[which.max(nchar(states))]
states[which.min(nchar(states))]
max(table(nchar(states)))
```


### Using ```grep()```

You can use the function ```grep()``` to know if there are any states containing the letter "z".
```{r}
# states containing the letter 'z'
grep(pattern = 'z', x = states)
```
In this case there is just one state (the third one) which corresponds to `r states[grep(pattern = 'z', x = states)]`

You can also use ```grep()``` with its argument ```value = TRUE``` to obtain the value of the matched pattern:
```{r}
# states containing the letter 'z'
grep(pattern = 'z', x = states, value = TRUE)
```

__Your turn__. Use ```grep()```---and maybe other functions---to write the commands that answer the following questions:

- How many states contain the letter ```i```?
- How many states contain the letter ```q```?
- How many states do not contain the letter ```a```?
- Which states contain the letter ```j```?
- Which states contain the letter ```x```?
- Which states are formed by two words?
- Which states start with ```W``` and end with a vowel?
- Which states start with ```W``` and end with a consonant?
- Which states contain at least three ```i``` (e.g. Illinois)?
- Which states contain five vowels (e.g. California)?
- Which states have three vowels next to each other (e.g. Hawaii)?

Tip: You can use ```grep()```'s argument ```ignore.case``` to ignore letters in lower or upper case.
```{r}
# your answers
length(grep("i", x = states))
length(grep("q", x = states))
length(grep("a", x = states, invert = TRUE))
states[grep("[jJ]", x = states)]
states[grep("[xX]", x = states)]
states[grep(". .", x = states)]
states[grep("^W.+[aeiou]$", x = states)]
states[grep("^W.+[^aeiou]$", x = states)]
states[grep("i.*i.*i.*", x = "iii", ignore.case = TRUE)]
setdiff(states[grep("[aeiou].*[aeiou].*[aeiou].*[aeiou].*[aeiou]", x = states, ignore.case = TRUE)], states[grep("[aeiou].*[aeiou].*[aeiou].*[aeiou].*[aeiou].*[aeiou]", x = states, ignore.case = TRUE)])
states[grep("[aeiou][aeiou][aeiou]", x= states, ignore.case = TRUE)]
```

-----

## Starts with ... 

Write a function ```starts_with()``` such that, given a character string and a single character, it determines whether the string starts with the provided character. 

```{r}
# starts_with
starts_with = function(charString, singleChar){
  singleChar = str_c("^", singleChar, "")
  return (any(grepl(singleChar, x = charString)))
}
```

Test it:
```{r, eval=TRUE}
starts_with("Hello", 'H')  # TRUE

starts_with("Good morning", 'H')  # FALSE
```

-----

## Ends with ... 

Now write a function ```ends_with()``` such that, given a character string and a single character, it determines whether the string ends with the provided character. 

```{r}
# ends_with
ends_with = function(charString, singleChar){
  singleChar = str_c(singleChar, "$")
  return (any(grepl(singleChar, x = charString)))
}
```

Test it:
```{r, eval=TRUE}
ends_with("Hello", 'o')  # TRUE

ends_with("Good morning", 'o')  # FALSE
```

-----

## Colors in Hexadecimal Notation

Write a function ```is_hex()``` that checks whether the input is a valid color in hexadecimal notation. Remember that a hex color starts with a hash ```#``` symbol followed by six hexadecimal digits: 0 to 9, and the first six letters A, B, C, D, E, F. Since R accepts hex-colors with lower case letters (a, b, c, d, e, f) your function should work with both upper and lower case letters.
```{r}
# is_hex()
is_hex = function(input){
  return (grepl("#[0-9a-z]{6}", x = tolower(input)))
}
```

Test it:
```{r, eval=TRUE}
is_hex("#FF00A7")  # TRUE
is_hex("#ff0000")  # TRUE
is_hex("#123456")  # TRUE
is_hex("#12Fb56")  # TRUE

is_hex("FF0000")   # FALSE
is_hex("#1234GF")  # FALSE
is_hex("#09892")   # FALSE
is_hex("blue")     # FALSE
```

-----

## Hexadecimal Colors with Transparency

Write a function ```is_hex_alpha()``` that determines whether the provided input is a hex color with alpha transparency. Remember that such a color has 8 hexadecimal digits instead of just 6.

```{r}
# is_hex_alph()
is_hex_alpha = function(input){
  return (nchar(input) == 9)
}
```

Test it:
```{r, eval=TRUE}
is_hex_alpha("#FF000078")  # TRUE

is_hex_alpha("#FF0000")   # FALSE
```

-----

## Splitting Characters

Create a function ```split_chars()``` that splits a character string into one single character elements.

```{r}
# split_chars()
split_chars = function(input){
  return (unlist(strsplit(input, "")))
}
```

Test it:
```{r, eval=TRUE}
split_chars('Go Bears!')

split_chars('Expecto Patronum')
```

Note that ```split_chars()``` returns the output in a single vector. Each element is a single character.

-----

## Number of Vowels

Create a function ```num_vowels()``` that returns the number of vowels of a character vector. In this case, the input is a vector in which each element is a single character.

```{r}
# num_vowels()
num_vowels = function(x){
  x = paste(gsub("[^aeiou]", "", x), collapse = "")
  a = length(unlist(str_extract_all(x, "a")))
  e = length(unlist(str_extract_all(x, "e")))
  i = length(unlist(str_extract_all(x, "i")))
  o = length(unlist(str_extract_all(x, "o")))
  u = length(unlist(str_extract_all(x, "u")))
  output = c(a,e,i,o,u)
  names(output) = c("a","e","i","o","u")
  return (output)
}
```

Test it:
```{r, eval=TRUE}
vec <- c('G', 'o', ' ', 'B', 'e', 'a', 'r', 's', '!')
num_vowels(vec)
```

Notice that the output is a numeric vector with five elements. Each element has the name of the corresponding vowel.

-----

## Counting Vowels

Use the functions ```split_chars()``` and ```num_vowels()``` to write a function ```count_vowels()``` that computes the number of vowels of a character string:

```{r}
# count_vowels()
count_vowels = function(x){
  return (num_vowels(split_chars(tolower(x))))
}
```

Test it:
```{r, eval=TRUE}
count_vowels("The quick brown fox jumps over the lazy dog")
```

Make sure that ```count_vowels()``` counts vowels in both lower and upper case letters:

```{r, eval=TRUE}
count_vowels("THE QUICK BROWN FOX JUMPS OVER THE LAZY DOG")
```

-----

## Number of Consonants

Write a function ```num_cons()``` that counts the number of consonants regardless of whether thery are in upper or lower case (just the number, not the counts of each letter)
```{r, eval=TRUE}
# num_cons()
num_cons = function(input){
  return (sum(table(str_extract_all(tolower(input), "[^aeiou[[:space:]]]"))))
}
```

Test it:
```{r, eval=TRUE}
fox <- "The quick brown fox jumps over the lazy dog"
num_cons(fox)
```

-----

## Reversing Characters

Write a function that reverses a string by characters

```{r}
# reverse_chars()
reverse_chars = function(x){
  return (paste(rev(unlist(str_split(x, ""))),collapse=""))
}
```

Test it:
```{r, eval=TRUE}
reverse_chars("gattaca")
reverse_chars("Lumox Maxima")
```

-----

## Reversing Sentences by Words

Write a function ```reverse_words()``` that reverses a string (i.e. a sentence) by words

```{r}
# reverse_words()
reverse_words = function(x){
  return (paste(rev(unlist(str_split(x, " "))), collapse = " ")) 
}
```

Test it:
```{r, eval=TRUE}
reverse_words("sentence! this reverse")
```

If the string is just one word then there's basically no reversing:
```{r, eval=TRUE}
reverse_words("string")
```

-----

